Topics in Algorithms

### 2018-01-10 project/paper.tex
Theorem apsidj

\section{Overview}
Given an undirected graph $G$ with non-negative edge weights, $m$ edges, $n$ vertices and some $k \in \mathbb{Z} \geq 2$, Thorup and Zwick showed that a $(2k-1)$-approximate distance oracle for $G$ of size $O(kn^{1+1/k})$ and $O(k)$ query time can be constructed in $O(kmn^{1/k})$ expected time \cite{Thorup2005}. \cite{Wulff2013} improves this query time to $\lg k$, while preserving stretch, space and preprocessing time. % Specifically $O( \min \left\{ kmn^{1/k}, \sqrt{k}m+kn^{1+c/\sqrt{k}} \right\})$ time for some constant $c$.

The query time reduction is made possible by observing that binary search can be performed on the iteration sequence of the $dist_k(u)$ algorithm described in \cite{Thorup2005}. This is done by precomputing a secondary data structures $\mathcal{T}_u$ for each $u \in V$ in $O(k)$ time. Each $\mathcal{T}_u$ contains information about which subsequences of $0, \hdots, k-1$ that make the modified $dist_k(u,i)$ algorithm (see Alg. \ref{alg:distk}) terminate while maintaining a  stretch of $(2k-1)$. Such subsequences are called $(u,v)$-feasible. Specifically, the subsequences in $T_u$ are \textit{canonical}. $(u,v)$-feasibility and canonicity is defined and discussed in 3.

% Assumptions about the reader

\section{Definitions}
The following definitions taken directly from \cite{Thorup2005} and \cite{Wulff2013} are used throughout this project and are listed here for brevity and coherency:
\begin{itemize}
\item $d_{G}(u,v)$: the shortest path distance between $u$ and $v$.
\item $A_i = \left\{ a \in A_{i-1} \; | \; a \text{ sampled with prob. } n^{-1/k} \text{ uniformly, i.i.d.} \right\}$
\item $(V = A_0) \supseteq A_1 \supseteq \hdots \supseteq (A_k = \emptyset)$
\item $p_i(u) = \min_{v \in A_i} \left\{ d_{G}(u,v) \right\}$
\item $B_u=\bigcup_{i=0}^{k-1}\left\{ v \in A_i - A_{i+1} \; | \; d_{G}(u,v) < d_{G}(u, p_{i+1}(u)) \right\}$, the bunch of $u$.
\item $d_G(u,p_k(u))=\infty$
\end{itemize}
Thorup and Zwick showed the property about bunches seen in Lemma \ref{lem:onedeltaperiteration}. The lemma is visualized in Fig. \ref{fig:lemma1vis}, and follows from the definitions of bunches. That $A_{k-1} \subset B_u$ and $p_{k-1}(v) \in B_u$ also readily follows from the definition of $A_{k}$ and $d_G(u,p_k(u))$ shown above:
\\
\begin{lem}
Let $u,v \in V$ be distinct vertices $0 \leq i < k-1$. If $p_i(v) \not\in B_{u}$ then $d_G(u, p_{i+1}(u)) \leq d_G(u, p_i(v))$. Furthermore $A_{k-1} \subset B_u$ and $p_{k-1}(v) \in B_u$. \label{lem:onedeltaperiteration}
\end{lem}
This lemma is readily used for bounding $d_G(u,v)$ by successively applying the triangle inequality at each iteration of $dist_k(u,v)$ to arrive at the $(2k-1)$-stretch bound in \cite{Thorup2005}.

\begin{figure}[!hbt]
\centering
    \includegraphics[width=0.5\textwidth]{include/tikz/lemma1.tikz}%
\caption{Visualization of Lemma \ref{lem:onedeltaperiteration}. The dotted green path corresponds to the circle defined by $u$ and the radii $d_G(p_i(u))$, which $p_{i+1}(v)$ is clearly outside. \label{fig:lemma1vis}}
\end{figure}

\section{Oracle with $O(\lg k)$ query time}
As aforementioned, binary search on the iteration sequence of the Thorup-Zwick oracle is performed to drive down the query time to $\lg k$. Starting with the iteration sequence $\mathcal{I}=1,\hdots,k-1$, the idea is to identify $r=O(\lg k)$ subsequences $(\mathcal{I}_1=\mathcal{I}) \supset \mathcal{I}_1 \supset \hdots \supset \mathcal{I}_r$, where for $2 \leq j \leq t$ we have $|\mathcal{I}_{j-1}| \geq 2|\mathcal{I}_j|$. We wish to maintain the invariant that $dist_k(u,v,i)$ applied to the first element of any subsequence $\mathcal{I}_j$ outputs the stretch $(2k-1)$ estimate in $|\mathcal{\mathcal{I}_j}|$ time.
\\
\begin{defi}
For vertices $u,v \in V$ an index $j \in \mathcal{I}$ is $(u,v)$-terminal if
\begin{enumerate}
\item $j = k-1$ or
\item $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ 
\end{enumerate}
We note that $j$ attains a $(u,v)$ terminal index, the algorithm terminates in the current or next iteration. \label{def:uvterm}
\end{defi}
The definition of $(u,v)$-terminality captures the essence of when $dist_k(u,v)$ terminates, i.e. when $p_j(u) \in B_v$. This happens either when 1) $j=k-1$, in which case, trivially, $p_{k-1}(u) \in B_v$ or 2) when $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ (in this case $dist_k(u,v)$ terminates either in iteration $j$ or $j+1$), see Alg. \ref{alg:distk}.
\\
\begin{defi}
A subsequence $\mathcal{I'}=i_1, \hdots, i_2$ of $\mathcal{I}$ is $(u,v)$-feasible if
\begin{enumerate}
\item $i_1$ is even and
\item $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$ and
\item $i_2$ is $(u,v)$-terminal
\end{enumerate}
\label{def:uvfeas}
\end{defi}
Furthermore, the definition of $(u,v)$-feasibility

\begin{Walgo}[ht]{20pt}
  \DontPrintSemicolon
  \SetKwRepeat{Do}{do}{while}
  \SetKwInOut{Input}{Input}
  \SetKwInOut{Output}{Output}
  \Indm
    \Indp
    \BlankLine
    $w \leftarrow p_i(u); j \leftarrow i$

@@

    \caption{$dist_k(u,v,i)$}
    \label{alg:distk}
\end{Walgo}

\begin{Walgo}[ht]{20pt}
  \DontPrintSemicolon
  \SetKwRepeat{Do}{do}{while}
  \SetKwInOut{Input}{Input}
  \SetKwInOut{Output}{Output}
  \Indm
    \Indp
    \BlankLine
    \If{$i_2-i_1 \leq \lg k$}{
    	\Return{$dist_k(u,v,i_1)$}
	}

    \textbf{let} $i$ be the middle even index $i_1,\hdots,i_2$

    \textbf{let} $i$ be the (precomputed) even index in $i_1,\hdots,i-2$ maximizing $\delta_j(u)$

    \If{$p_j(u) \not\in B_v$ \textbf{and} $p_{j+1} \not\in B_u$}{
    	\Return{$bdist_k(u,v,i,i_2)$}
    }
    \Else{
    	\Return{$bdist_k(u,v,i_1,j)$}
    }
    \caption{$bdist_k(u,v,i_1,i_2)$}
    \label{alg:bdistk}
\end{Walgo}

@@

Furthermore $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$. When $p_j(u) = w \not\in B_v$
we have $d_G(v,p_{j+1}(v)) \leq d_G(v,p_{j}(u)) \leq d_G(v,u) + d_G(u,p_j(u))$ by Lemma \ref{lem:onedeltaperiteration} and the triangle inequality. Hence each iteration of $dist_k(u,v,i)$ increases $d_G(u,w)$ by at most $d_G(u,v)$. At termination we have
\begin{align*}
d_G(u,w) + d_G(w,v) &\leq 2d_G(u,w) + d_G(u,v) \\
&\leq 2(i_1 \cdot d_G(u,v) + (i_2 - i_1)d_G(u,v)) + d_G(u,v) \\
&\leq 2 \cdot i_2 \cdot d_G(u,v) + d_G(u,v) \\
&\leq (2 \cdot i_2 + 1)d_G(u,v) \\
&\leq (2 (k-1) + 1)d_G(u,v) \\
&\leq (2k-1) d_G(u,v)
\end{align*}
\end{lem}
% Inkluder illustration - forklarende tekst "By lemma 2.1 and the triangle inequality"

@@

\begin{figure}[h]
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=3cm},
  level 2/.style={sibling distance=1.5cm}]
  \node[color=red] {$I'$}
    child[color=red] {node {$[0, ..., i]$}
    child[color=black,thick] {node[color=cyan] {$I^j$}}
      child[dashed] {node {$p(l_1)$}
      	child[solid, left=0.1cm, color=black] {node[color=cyan] {$l_1=[i_1, i_1+2]$}}
      	child[solid,color=black] {node[color=cyan] {$[i_1+2,i_1+4]$}}
      }
    }
    child[right=0.1cm, color=red] {node {$[i,...,k-1]$}
    child[color=black] {node[color=cyan] {$I^k$} 
    						child[dotted, color=black] {node {...}}
    						child[dotted, color=black,left=0.1cm] {node {...}}
    					}
      child[dashed] {{node {$p(l_2)$}
       	child[solid,color=black] {node[color=cyan] {$[i_2-4,i_2-2]$}}
      	child[right=1cm,solid,color=black] {node[color=cyan] {$l_2=[i_2-2, i_2]$}}
    }}};

@@

\begin{thm}
For an integer $k\geq 2$, a $(2k - 1)$-approximate distance oracle of G of size $O(kn^{1+1/k})$ and $O(\log k)$ query time can be constructed in $O\left( \min \left\lbrace kmn^{1+\frac{1}{k}}, \sqrt{k}m + kn^{1+c/k}\right\rbrace\right)$ time for some constant $c$.
\end{thm}
{\it Proof.} For variable $k\geq 3$, for each $u$ the bunch $B_u$ in computed in $O(kmn^{2})$ using \cite{Thorup2005}. The remaining preprocessing for $u$ is determining the necessary subsequences of $\mathcal{I}$ and the $j$ that maximized $\delta_j(u)$ for each subsequence.
The necessary subsequences are built

@@

The recursion tree for $bdist_k$ for $u$ will have $\theta(k/\log k)$ children and $\theta(k/\log k)$ non-child nodes. Computing $j$ for each subsequence in $T$ will take $O(\log k)$ time per node, so total preprocessing time for $u$ is $O(k)$.  
\bibliographystyle{plain}
\bibliography{bibliography}

>>>>

\section{Overview}
Given an undirected graph $G$ with non-negative edge weights, $m$ edges, $n$ vertices and some $k \in \mathbb{Z} \geq 2$, Thorup and Zwick showed that a $(2k-1)$-approximate distance oracle for $G$ of size $O(kn^{1+1/k})$ and $O(k)$ query time can be constructed in $kmn^{1/k}$ expected time \cite{Thorup2005}. \cite{Wulff2013} improves this query time to $\lg k$, while preserving stretch, space and preprocessing time. % Specifically $O( \min \left\{ kmn^{1/k}, \sqrt{k}m+kn^{1+c/\sqrt{k}} \right\})$ time for some constant $c$.

The reduction in query time is accomplished by performing binary search on the bunch structures described in \cite{Thorup2005} by uniquely determining at which level the bunch 
% Assumptions about the reader

% Reference til Mikkel/Zwick paper - hans query time og 'vores' forbedrede query time - forbedring af konstruktion. Forbedring i querytime, uden at det påvirker space/konstruktionstid.

% Motivation

% Girth Conjecture?

\section{Definitions}
The following definitions are taken directly from \cite{Thorup2005} and \cite{Wulff2013} and are used throughout this project.
\begin{itemize}
\item $d_{G}(u,v)$: the shortest path distance between $u$ and $v$.
\item $A_i = \left\{ a \in A_{i-1} \; | \; a \text{ sampled with prob. } n^{-1/k} \text{ uniformly, i.i.d.} \right\}$
\item $V = A_0 \supseteq A_1 \supseteq \hdots \supseteq A_k = \emptyset$
\item $p_i(u) = \min_{v \in A_i} \left\{ d_{G}(u,v) \right\}$
\item $B_u=\bigcup_{i=0}^{k-1}\left\{ v \in A_i - A_{i+1} \; | \; d_{G}(u,v) < d_{G}(u, p_{i+1}(u)) \right\}$, the bunch of $u$.
\item $d_G(u,p_k(u))=\infty$
\end{itemize}
Thorup and Zwick showed that
\\
\begin{lem}
Let $u,v \in V$ be distinct vertices $0 \leq i < k-1$. If $p_i(v) \not\in B_{u}$ then $d_G(u, p_{i+1}(u)) \leq d_G(u, p_i(v))$. Furthermore $A_{k-1} \subset B_u$ and $p_{k-1}(v) \in B_u$. \label{lem:onedeltaperiteration}
\end{lem}

\section{Oracle with $O(\lg k)$ query time}
\begin{Walgo}[ht]{20pt}
  \DontPrintSemicolon
  \SetKwRepeat{Do}{do}{while}
  \SetKwInOut{Input}{Input}
  \SetKwInOut{Output}{Output}
  \Indm
    \Indp
    \BlankLine
    $w \leftarrow p_i(u); j \leftarrow i$

@@

    \caption{$dist_k(u,v,i)$}
    \label{alg:0.1}
\end{Walgo}

\begin{defi}
For vertices $u,v \in V$ an index $j \in \mathcal{I}$ is $(u,v)$-terminal if
\begin{enumerate}
\item $j = k-1$ or
\item $j < k-1$ is even and $p_j(u) \in B_v$ or $p_{j+1}(v) \in B_u$ 
\end{enumerate}
We note that $j$ attains a $(u,v)$ terminal index, the algorithm terminates in the current or next iteration.
\end{defi}

\begin{defi}
A subsequence $\mathcal{I'}=i_1, \hdots, i_2$ of $\mathcal{I}$ is $(u,v)$-feasible if
\begin{enumerate}
\item $i_1$ is even and
\item $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$ and
\item $i_2$ is $(u,v)$-terminal
\end{enumerate}
\end{defi}

@@

Furthermore $d_{G}(p_{i_1}(u),u) \leq i_1 \cdot d_G(u,v)$. When $p_j(u) = w \not\in B_v$
we have $d_G(v,p_{j+1}(v)) \leq d_G(v,p_{j}(u)) \leq d_G(v,u) + d_G(u,p_j(u))$ by Lemma \ref{lem:onedeltaperiteration} and the triangle inequality, as visualized in Fig. \ref{fig:j1}. Hence each iteration of $dist_k(u,v,i)$ increases $d_G(u,w)$ by at most $d_G(u,v)$. At termination we have
\begin{align*}
d_G(u,w) + d_G(w,v) &\leq 2d_G(u,w) + d_G(u,v) \\
&\leq 2(i_1 \cdot d_G(u,v) + (i_2 - i_1)d_G(u,v)) + d_G(u,v) \\
&\leq 2 \cdot i_2 \cdot d_G(u,v) + d_G(u,v) \\
&\leq (2 \cdot i_2 + 1)d_G(u,v) \\
&\leq (2 (k-1) + 1)d_G(u,v) \\
&\leq (2k-1) d_G(u,v)
\end{align*}
\end{lem}
% Inkluder illustration - forklarende tekst "By lemma 2.1 and the triangle inequality"

@@

\begin{figure}[H]
\begin{tikzpicture}[level distance=1.5cm,
  level 1/.style={sibling distance=3cm},
  level 2/.style={sibling distance=1.5cm}]
  \node[color=red] {$I'$}
    child[color=red] {node {$[0, ..., i]$}
    child[color=black,thick] {node[color=cyan] {$I^j$}}
      child[dashed] {node {$p(l_1)$}
      	child[solid, left=0.1cm, color=black] {node[color=cyan] {$l_1=[i_1, i_1+2]$}}
      	child[solid,color=black] {node[color=cyan] {$[i_1+2,i_1+4]$}}
      }
    }
    child[right=0.1cm, color=red] {node {$[i,...,k-1]$}
    child[color=black] {node[color=cyan] {$I^k$} 
    						child[dotted, color=black] {node {...}}
    						child[dotted, color=black,left=0.1cm] {node {...}}
    					}
      child[dashed] {{node {$p(l_2)$}
       	child[solid,color=black] {node[color=cyan] {$[i_2-4,i_2-2]$}}
      	child[right=1cm,solid,color=black] {node[color=cyan] {$l_2=[i_2-2, i_2]$}}
    }}};

@@

\begin{thm}
For an integer $k\geq 2$, a $(2k - 1)$-approximate distance oracle of G of size $O(kn^{1+1/k})$ and $O(\log k)$ query time can be constructed in $O\left( \min \left\lbrace kmn^{1+\frac{1}{k}}, \sqrt{k}m + kn^{1+c/k}\right\rbrace\right)$ time for some constant $c$.

\paragraph{Proof.} For variable $k\geq 3$, for each $u$ the bunch $B_u$ in computed in $O(kmn^{2})$ using \cite{Thorup2005}. The remaining preprocessing for $u$ is determining the necessary subsequences of $\mathcal{I}$ and the $j$ that maximized $\delta_j(u)$ for each subsequence.
The necessary subsequences are built by

@@

The recursion tree for $bdist_k$ for $u$ will have $\theta(k/\log k)$ children and $\theta(k/\log k)$ non-child nodes. Computing $j$ for each subsequence in $T$ will take $O(\log k)$ time per node, so total preprocessing time for $u$ is $O(k)$.  
\end{thm}
\bibliographystyle{plain}
\bibliography{bibliography}
